\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2019-01-13 archive.tex
more writing

We view these wake sleep/cycles as an approximate inference algorithm that starts with a collection of tasks, written $X$, and infers both a program for each task, as well
as a distribution over programs encoded by a DSL, written
$\mathcal{D}$.  We equip $\mathcal{D}$ with a learned weight vector
$\theta$, and together $(\mathcal{D},\theta)$ define a generative
model over programs (Appendix~\ref{sampleProgram}).  We perform
maximum a posteriori (MAP) inference of the DSL and programs given
$X$.  Writing $J$ for the joint probability of $(\mathcal{D},\theta)$
and $X$, we seek the generative model $\mathcal{D}^*,\theta^*$ solving:
\begin{align}\label{intractableObjectives}
\nonumber  J(\mathcal{D},\theta)\triangleq \probability[\mathcal{D},\theta]&\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\\
  \mathcal{D}^* = \argmax_{\mathcal{D}}\int J(\mathcal{D},\theta)\;\mathrm{d}\theta &\qquad
  \theta^* =\argmax_\theta J(\mathcal{D}^*,\theta)
\end{align}
%% where $\probability[x|p]$ scores the likelihood of a task
%% $x\in X$ given a program $p$.\footnote{For example, for list
%%   processing, the likelihood is 1 if the program predicts the observed
%%   outputs on the observed inputs, and 0 otherwise; when learning a
%%   generative model or probabilistic program, the likelihood is the
%%   probability of the program sampling the observation.}
Evaluating Eq.~\ref{intractableObjectives}
entails marginalizing over the infinite set of all programs -- which is impossible.
We make a particle-based approximation to Eq.~\ref{intractableObjectives}
and instead marginalize over a finite \textbf{beam} of programs,
with one beam per task, collectively written $\left\{\mathcal{B}_x \right\}_{x\in X}$.
This particle-based approximation is written $\lowerBound (\mathcal{D},\theta,\left\{\mathcal{B}_x \right\})$
and acts as a lower bound on the joint:
\begin{align}
 J(\mathcal{D},\theta)\geq \lowerBound  (\mathcal{D},\theta,\left\{\mathcal{B}_x \right\})\triangleq\probability[\mathcal{D},\theta]\prod_{x\in X} \sum_{p\in \mathcal{B}_x} \probability[x|p]\probability[p|\mathcal{D},\theta]\text{, where $|\mathcal{B}_x|$ is small}
\end{align}
Wake and sleep cycles correspond to alternate maximization of $\lowerBound$ w.r.t. $\left\{\mathcal{B}_x \right\}_{x\in X}$  (\textbf{Wake})
and $(\mathcal{D},\theta)$ (\textbf{Consolidation}):
\\\noindent \textbf{Wake: Maxing $\lowerBound$ w.r.t.\ the beams.} Here $(\mathcal{D},\theta)$ is fixed and we
want to find new programs to add to  the beams so that $\lowerBound$ increases the most.
$\lowerBound$ most increases by finding programs where $\probability[x|p]\probability[p|\mathcal{D},\theta]\propto\probability[p|x,\mathcal{D},\theta]$ 
is large (i.e., programs with high posterior probability).
%Here we use a simple and generic enumerative program synthesis algorithm.
%% which we can accomplish by adding new programs to the beams means searching for new programs $p$ for task $x$
%% where  is large.
\\\noindent \textbf{Sleep (Consolidation): Maxing $ \lowerBound$ w.r.t.\ the DSL.} Here $\left\{\mathcal{B}_x \right\}_{x\in X}$ is held fixed, and so we can evaluate $\lowerBound$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $\int \lowerBound \;\mathrm{d}\theta$,
and then updating $\theta$ to $\argmax_\theta \lowerBound(\mathcal{D},\theta,\left\{\mathcal{B}_x \right\})$.

Searching for programs is hard because
of the large combinatorial search space. We ease this difficulty by training a neural recognition model, $Q(\cdot |\cdot )$,
during the \textbf{Dreaming} phase: $Q$ is trained to approximate the
posterior over programs, $Q(p|x)\approx \probability[p|x,\mathcal{D}]\propto\probability[x|p]\probability[p|\mathcal{D}]$.
  Thus training the neural network amortizes the cost of finding programs with high posterior probability.

@@

Intuitively, this 3-phase inference procedure can work in practice because each of the 3 phases bootstraps off of the others. As the DSL grows and as the recognition model becomes more accurate, waking becomes more effective, allowing the agent to solve more tasks; when we solve more tasks during waking,
the Consolidation cycle has more data from which to learn the DSL; and, because the recognition model is trained on both
samples from the DSL and programs found during waking,
the recognition model gets both more data, and higher-quality data, whenever the DSL improves and whenever we discover more successful programs.
\begin{figure}
  \begin{tikzpicture}[line width=0.4mm]
    \draw[fill=white!5!white] (-1,1) -- (13,1) -- (13,-4) -- (-1,-4) -- (-1,1);
    \node at (5.5,1.25) {\textsc{\textbf{Wake}}};

@@

\subsection{Wake: Solving tasks}\label{explorationSection}

@@

Why enumerate, when the program synthesis community has invented many
sophisticated algorithms that search for programs?~\cite{solar2008program,schkufza2013stochastic,feser2015synthesizing,osera2015type,polozov2015flashmeta}.
We have two reasons:
(1) A key point of our work is that learning the DSL, along with a neural recognition model, can make program induction tractable, even if the search algorithm is very simple.
(2) Enumeration is a general approach that can be applied to any program induction problem. Many of these more sophisticated approaches require special conditions on
the space of  programs.
\begin{comment}
  \begin{figure}
    \begin{tikzpicture}[every node/.style={rounded corners,thick}]
      \node(x)[draw,align=center] at (0,0) {Task $x$\\e.g., input/outputs};
      \node(e)[draw, align=center] at ([xshift=3cm]x.east) {Domain-specific encoder\\
        e.g., RNN or CNN};
      %    \node(m)[draw,align = center] at ([xshift=1.1cm]e.east) {MLP};%\\Output: $(|\mathcal{D}| + 2)\times(|\mathcal{D}| + 1)\times A$};
      \node[draw,align = center](s) at ([xshift=2cm]e.east) {Enumerate\\(Algorithm~\ref{recognitionSample})};
      \node(f)[draw,align = center] at ([xshift=2cm]s.east) {Beam\\$\mathcal{B}_x$};

@@

The DSL offers a set of abstractions that allow an agent to concisely
express solutions to the tasks at hand. We automatically discover
these new abstractions by combining two ideas. First, we build on
techniques from the programming languages community to develop a new
algorithm for automatically refractoring programs, where this
refactoring exposes common reused subexpressions found across the
programs found during waking.  Second, we use this automatic refactoring
process to search for DSLs that maximally compress these programs by
incorporating reused subexpressions into the DSL.

Mathematically this compression takes the form of
finding the DSL maximizing $\int \lowerBound\;\mathrm{d}\theta$ (Sec.~\ref{overviewSection}).
We replace this marginal with an AIC approximation~\cite{akaike1998information}
and minimize the following expression,
which can be interpreted as a kind of compression:
\begin{equation}
\underbrace{-\log \probability[\mathcal{D}] + \min_{\theta}\Bigg(-\log
\probability[\theta|\mathcal{D}] + \|\theta\|_0}_{\text{Description length of }(\mathcal{D},\theta)} +\sum_{x\in
  X}\underbrace{-\log \sum_{p\in
  \mathcal{B}_x}\probability[x|p]\probability[p|\mathcal{D},\theta]}_{\text{Description length of programs for task }x}\Bigg)
\label{AIC}
  \end{equation}
At a high level, our approach is to search locally through the space 
of DSLs, proposing small changes to $\mathcal{D}$ until Eq.~\ref{AIC}
fails to decrease.  These small changes consist of introducing new
candidate $\lambda$-expressions into the DSL.

However, there is a snag with this simple approach:
whenever we add a new expression $e$ to the DSL,
the programs found during waking ($\left\{\mathcal{B}_x \right\}$)
are not written in terms of $e$.
Concretely, imagine we wanted to discover a new DSL procedure for doubling numbers,
after having found the programs
\code{(cons (+ 9 9) nil)} and \code{($\lambda$ (x) (+ (car x) (car x)))}.
As human programmers,
we can look at these pieces of code and recognize that,
if we define a new procedure called \code{double},
defined as \code{($\lambda$ (x) (+ x x))},
then we can rewrite the original programs as
\code{(cons (double 9) nil)} and \code{($\lambda$ (x) (double (car x)))}.
This process is a kind of refactoring
where a new subroutine is defined (\code{double})
and the old programs rewritten
in terms of the new subroutine.
Figure~\ref{mapFactor}A
diagrams this refactoring
process for a more complicated
setting,
where the agent must rediscover the higher-order function \code{map}
starting from the basics of Lisp and the Y-combinator.

We refine our objective to compress
\emph{refactorings} of the programs found during waking, minimizing
\begin{equation}
        -\log \probability[\mathcal{D}] + 
\min_{\theta}\Bigg(-\log \probability[\theta|\mathcal{D}] + \|\theta\|_0 +\sum_{x\in X} - \log \underbrace{\sum_{\substack{p:\\\exists p'\in \mathcal{B}_x: p\manyReduce p'}}}_{\text{Refactors }\mathcal{B}_x}\probability[x|p]\probability[p|\mathcal{D},\theta]\Bigg)
\label{factorObjective}
\end{equation}
where $p\manyReduce p'$ is the standard notation for ``expression $p$
evaluates to $p'$ by the rules of $\lambda$-calculus''~\cite{pierce}.
Equation~\ref{factorObjective} captures the idea that we want to add
new components to the DSL while jointly refactoring our old programs
in terms of these new components.  But this joint optimization is
intractable, because there are infinitely many ways of refactoring a
program.  To make refactoring tractable we first limit the degree to
which a piece of code can be refactored: rather than consider every
refactoring, we bound the number of $\lambda$-calculus evaluation
steps separating a refactoring from its original program.  Now the
number of refactoring this is finite but astronomically large: for the
example in Figure~\ref{mapFactor} there are approximately $10^{14}$
possible refactorings -- a quantity that grows exponentially both as a
function of program size and a function of the bound on evaluation
steps. How can we tame this combinatorial explosion?

A structure called a \textbf{version space
  algebra}~\cite{lau2001programming,mitchell1977version,polozov2015flashmeta}
neatly resolves this exponential growth.  A version space is a
tree-shaped data structure that compactly represents a large set of
programs and supports efficient set operations like union,
intersection, and membership checking.  In
Appendix~\ref{appendixVersion}, we give a dynamic program that takes
as input a program $p$ and then outputs a version space containing the
refactorings of $p$. Figure~\ref{mapFactor}B diagrams a subtree of a
version space containing refactorings of a small program. Our
technique is substantially more efficient than explicitly representing
the space of possible refactorings: for the example in
Figure~\ref{mapFactor}A, we represent the space of refactorings using
a version space with $10^6$ nodes, which encodes $10^{14}$
refactorings. Appendix~\ref{appendixCompression} specifies how we combine this probabilistic and symbolic machinery to update the DSL.

@@

  \end{tikzpicture}
  \caption{DSL learning as code refactoring. \textbf{Panel A:} For each task we discover programs during waking, then refactor the code from those programs to expose common subprograms (highlighted in \orange{orange}). Common subprograms are incorporated into the DSL when they increase a Bayesian objective. Intuitively, these new DSL components best compress the programs found during waking. \textbf{Panel B:} \# of possible refactorings grows exponentially with program size, so we represent refactorings using version spaces, which augment syntax trees with a \emph{union} operator whose children are themselves version spaces. Right graph: version spaces are exponentially more efficient than explicitly constructing set of refactorings. In this graph, refactored programs are of the form $1+1+\cdots  + 1$.}\label{mapFactor}
\end{figure*}

@@

Formally, the recognition model $Q(p|x)$ should approximate the posterior
$\probability[p|\mathcal{D},\theta,x]$.
We can either train $Q$ to perform full posterior inference by minimizing the expected KL-divergence, $  \expect\left[\text{KL}\left(\probability[p|x,\mathcal{D},\theta]\|Q(p|x) \right) \right]$,
or we can train $Q$ to perform MAP inference
by maximizing $\expect\left[\max_{p\text{ maxing }\probability[\cdot |x,\mathcal{D},\theta]} \log Q(p|x) \right]$,
where in both cases the expectation is taken over tasks. Taking this expectation over the empirical distribution of tasks trains $Q$ on replays; taking it over samples from the generative model trains $Q$ on fantasies.
We define a pair of alternative objectives for the recognition model,
$\mathcal{L}^{\text{posterior}}$ and $\mathcal{L}^{\text{MAP}}$,
which either train $Q$ to perform full posterior inference or MAP inference, respectively.
These objectives combine replays and fantasies:
\begin{align*}
  \mathcal{L}^{\text{posterior}} &= \mathcal{L}_{\text{Replay}}^{\text{posterior}} + \mathcal{L}_{\text{Fantasy}}^{\text{posterior}}&
  \mathcal{L}^{\text{MAP}} &= \mathcal{L}_{\text{Replay}}^{\text{MAP}} + \mathcal{L}_{\text{Fantasy}}^{\text{MAP}}\\
  \mathcal{L}_{\text{Replay}}^{\text{posterior}}& = \expect_{x\sim X}\left[\sum_{p\in \mathcal{B}_x}
    \frac{\probability\left[x,p|\mathcal{D},\theta \right]\log Q(p|x)}{\sum_{p'\in \mathcal{B}_x}\probability\left[x,p'|\mathcal{D},\theta \right]}\right] &
  \mathcal{L}_{\text{Replay}}^{\text{MAP}}& = \expect_{x\sim X}\left[\max_{\substack{p\in \mathcal{B}_x\\p\text{ maxing }\probability[\cdot |x,\mathcal{D},\theta]}} \log Q(p|x) \right]  \\
  \mathcal{L}_{\text{Fantasy}}^{\text{posterior}} &= \expect_{(p,x)\sim(\mathcal{D},\theta) }\left[\log Q(p|x)\right]&
  \mathcal{L}_{\text{Fantasy}}^{\text{MAP}} &= \expect_{x\sim(\mathcal{D},\theta) }\left[\max_{\substack{p\\p\text{ maxing }\probability[\cdot |x,\mathcal{D},\theta]}}\log Q(p)\right]
\end{align*}
We maximize $\mathcal{L}^{\text{MAP}}$ rather than
$\mathcal{L}^{\text{posterior}}$ for two reasons:
$\mathcal{L}^{\text{MAP}}$ prioritizes the shortest program solving a
task, thus more strongly accelerating enumerative search during waking;
and, combined with our parameterization of $Q$, described next, we
will show that $\mathcal{L}^{\text{MAP}}$ forces the recognition model
to break symmetries in the space of programs
(Sec.~\ref{symmetricSection}).

@@

We adopt a middle ground between these two extremes.  Our recognition
model predicts a distribution over primitives in the DSL,
conditioned on the local context in the syntax tree of the
program. When predicting the next node to add to the syntax tree of a program,
the recognition model
conditions on the parent node, as well as
which argument is being generated.
This is a kind of `bigram' model over trees,
where the bigrams
take the form of (parent, child, argument index).
Figure~\ref{symmetry} (left) diagrams this generative process and Algorithm~\ref{recognitionSample}
specifies a sampling procedure for $Q(\cdot |x)$.
This parameterization
has three main advantages:
(1) it supports fast enumeration
and sampling of programs,
because the
recognition model
only needs to run once for each task;
(2) it allows the recognition model to provide fine-grained
information about the structure of the target program;
and (3)
training this recognition model
causes it to learn to break symmetries in the space of programs,
described next.

@@

\subsection{Programs that manipulate sequences}\label{sequences}
We first apply \system to two classic benchmark domains: list
processing and text editing. In both cases we solve tasks specified by
a few input/output examples, using a recurrent neural network for the
recognition model and starting with a generic functional programming
basis: \code{foldr}, \code{unfold}, \code{if}, \code{map},
\code{length}, \code{index}, \code{=}, \code{+}, \code{-}, \code{0},
\code{1}, \code{cons}, \code{car}, \code{cdr}, \code{nil}, and
\code{is-nil}.

>>>>

We view these wake sleep/cycles as an approximate inference algorithm that starts with a collection of tasks, written $X$, and infers both a program for each task, as well
as a distribution over programs encoded by a DSL, written
$\mathcal{D}$.  We equip $\mathcal{D}$ with a learned weight vector
$\theta$, and together $(\mathcal{D},\theta)$ define a generative
model over programs (Appendix~\ref{sampleProgram}).  We perform
maximum a posteriori (MAP) inference of the DSL and programs given
$X$.  Writing $J$ for the joint probability of $(\mathcal{D},\theta)$
and $X$, we seek the generative model $\mathcal{D}^*,\theta^*$ solving:
\begin{align}\label{intractableObjectives}
\nonumber  J(\mathcal{D},\theta)\triangleq \probability[\mathcal{D},\theta]&\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\\
  \mathcal{D}^* = \argmax_{\mathcal{D}}\int J(\mathcal{D},\theta)\;\mathrm{d}\theta &\qquad
  \theta^* =\argmax_\theta J(\mathcal{D}^*,\theta)
\end{align}
%% where $\probability[x|p]$ scores the likelihood of a task
%% $x\in X$ given a program $p$.\footnote{For example, for list
%%   processing, the likelihood is 1 if the program predicts the observed
%%   outputs on the observed inputs, and 0 otherwise; when learning a
%%   generative model or probabilistic program, the likelihood is the
%%   probability of the program sampling the observation.}
Evaluating Eq.~\ref{intractableObjectives}
entails marginalizing over the infinite set of all programs -- which is impossible.
We make a particle-based approximation to Eq.~\ref{intractableObjectives}
and instead marginalize over a finite \textbf{beam} of programs,
with one beam per task, collectively written $\left\{\mathcal{B}_x \right\}_{x\in X}$.
This particle-based approximation is written $\lowerBound (\mathcal{D},\theta,\left\{\mathcal{B}_x \right\})$
and acts as a lower bound on the joint density:
\begin{align}
 J(\mathcal{D},\theta)\geq \lowerBound  (\mathcal{D},\theta,\left\{\mathcal{B}_x \right\})\triangleq\probability[\mathcal{D},\theta]\prod_{x\in X} \sum_{p\in \mathcal{B}_x} \probability[x|p]\probability[p|\mathcal{D},\theta]\text{, where $|\mathcal{B}_x|$ is small}
\end{align}
Wake and sleep cycles correspond to alternate maximization of $\lowerBound$ w.r.t. $\left\{\mathcal{B}_x \right\}_{x\in X}$  (\textbf{Wake})
and $(\mathcal{D},\theta)$ (\textbf{Consolidation}):
\\\noindent \textbf{Wake: Maxing $\lowerBound$ w.r.t.\ the beams.} Here $(\mathcal{D},\theta)$ is fixed and we
want to find new programs to add to  the beams so that $\lowerBound$ increases the most.
$\lowerBound$ most increases by finding programs where $\probability[x|p]\probability[p|\mathcal{D},\theta]\propto\probability[p|x,\mathcal{D},\theta]$ 
is large, i.e., programs with high posterior probability,
or programs which are a priori likely and which solve the task.
We find programs solving a task by
enumerating programs
from the DSL in decreasing order of their probability
under the recognition model,
and then check if a program $p$ assigns positive probability to a task
($\probability[x|p] > 0$); if so, we incorporate $p$ into the beam
$\mathcal{B}_x$. We represent programs as polymorphicly typed $\lambda$-calculus expressions,
an expressive formalism
including conditionals,
variables,
higher-order recursive functions,
and the ability to define new functions.
\\\noindent \textbf{Sleep (Consolidation): Maxing $ \lowerBound$ w.r.t.\ the DSL.} Here $\left\{\mathcal{B}_x \right\}_{x\in X}$ is held fixed and the problem is to search the discrete space of DSLs and find one maximizing $\int \lowerBound \;\mathrm{d}\theta$,
and then update $\theta$ to $\argmax_\theta \lowerBound(\mathcal{D},\theta,\left\{\mathcal{B}_x \right\})$.

Finding programs solving tasks is difficult because of the infinitely
large, combinatorial search landscape. We ease this difficulty by
training a neural recognition model, $Q(p |x )$, during the
\textbf{Dreaming} phase: $Q$ is trained to approximate the posterior
over programs, $Q(p|x)\approx
\probability[p|x,\mathcal{D}]\propto\probability[x|p]\probability[p|\mathcal{D}]$.
Thus training the neural network amortizes the cost of finding
programs with high posterior probability.

@@

This 3-phase inference procedure works through two distinct kinds of bootstrapping.
During each sleep cycle the next DSL
bootstraps off the primitives learned
during earlier cycles,
growing an increasingly deep learned program representation (`deep program learning').
In tandem the DSL and recognition model
bootstrap each other:
a more finely tuned
DSL yields
richer dreams for the recognition model to learn from,
while a more accurate recognition model
solves more tasks during waking which then feed into the next DSL.
\begin{figure}
  \begin{tikzpicture}[line width=0.4mm]
    \draw[fill=white!5!white] (-1,1) -- (13,1) -- (13,-4) -- (-1,-4) -- (-1,1);
    \node at (5.5,1.25) {\textsc{\textbf{Wake}}};

@@

\begin{comment}
  \subsection{Wake: Solving tasks}\label{explorationSection}

@@

Why enumerate, when the program synthesis community has invented many
sophisticated algorithms that search for programs?~\cite{solar2008program,schkufza2013stochastic,feser2015synthesizing,osera2015type,polozov2015flashmeta}.
We have two reasons:
(1) A key point of our work is that learning the DSL, along with a neural recognition model, can make program induction tractable, even if the search algorithm is very simple.
(2) Enumeration is a general approach that can be applied to any program induction problem. Many of these more sophisticated approaches require special conditions on
the space of  programs.
\end{comment}
\begin{comment}
  \begin{figure}
    \begin{tikzpicture}[every node/.style={rounded corners,thick}]
      \node(x)[draw,align=center] at (0,0) {Task $x$\\e.g., input/outputs};
      \node(e)[draw, align=center] at ([xshift=3cm]x.east) {Domain-specific encoder\\
        e.g., RNN or CNN};
      %    \node(m)[draw,align = center] at ([xshift=1.1cm]e.east) {MLP};%\\Output: $(|\mathcal{D}| + 2)\times(|\mathcal{D}| + 1)\times A$};
      \node[draw,align = center](s) at ([xshift=2cm]e.east) {Enumerate\\(Algorithm~\ref{recognitionSample})};
      \node(f)[draw,align = center] at ([xshift=2cm]s.east) {Beam\\$\mathcal{B}_x$};

@@

The DSL offers a set of abstractions that allow an agent to concisely
express solutions to the tasks at hand. We automatically discover
these new abstractions by combining two ideas. First, we build on
techniques from the programming languages community to develop a new
algorithm for automatically refractoring programs, where this
refactoring exposes common reused subexpressions across the
programs found during waking.  Second, we use this automatic refactoring
process to search for DSLs that maximally compress these programs by
incorporating reused subexpressions into the DSL.

Mathematically this compression takes the form of
finding the DSL maximizing $\int \lowerBound\;\mathrm{d}\theta$ (Sec.~\ref{overviewSection}).
We replace this marginal with an AIC approximation~\cite{akaike1998information}
and marginalize over refactorings of programs in the beams, minimizing the following expression,
which can be interpreted as a kind of compression:
\begin{equation}
\underbrace{-\log \probability[\mathcal{D}] + \min_{\theta}\Bigg(-\log
\probability[\theta|\mathcal{D}] + \|\theta\|_0}_{\text{Description length of }(\mathcal{D},\theta)} +\sum_{x\in
  X}\underbrace{-\log \sum_{\substack{p:\\
      \exists p'\in
  \mathcal{B}_x: p\text{ refactors into }p'}}\probability[x|p]\probability[p|\mathcal{D},\theta]}_{\text{Description length of programs for task }x}\Bigg)
\label{AIC}
\end{equation}
But a program has infinitely many possible refactorings, rendering
Eq.~\ref{AIC} intractable.  Rather than consider every refactoring we
bound the number of $\lambda$-calculus evaluation steps separating a
program from its refactoring.  Now the number of refactorings is
finite but astronomically large: Figure~\ref{mapFactor}A diagrams a
problem where the agent rediscovers the higher-order function
\code{map} starting from the basics of Lisp and the Y-combinator, but
where there are approximately $10^{14}$ possible refactorings -- a quantity
that grows exponentially both as a function of program size and a
function of the bound on evaluation steps. How can we tame this
combinatorial explosion?

A structure called a \textbf{version space
  algebra}~\cite{lau2001programming,mitchell1977version,polozov2015flashmeta}
neatly resolves this exponential growth.  A version space is a
tree-shaped data structure that compactly represents a large set of
programs and supports efficient set operations like union,
intersection, and membership checking.  In
Appendix~\ref{appendixVersion}, we give a dynamic program that takes
as input a program and then outputs a version space containing its
refactorings. Figure~\ref{mapFactor}B diagrams a subtree of a
version space containing refactorings of a small program. Our
technique is substantially more efficient than explicitly representing
the space of possible refactorings: for the example in
Figure~\ref{mapFactor}A, we represent the space of refactorings using
a version space with $10^6$ nodes, which encodes $10^{14}$
refactorings. Appendix~\ref{appendixCompression} specifies how we combine this probabilistic and symbolic machinery to update the DSL.
At a high level, our approach is to search locally through the space 
of DSLs, proposing small changes until Eq.~\ref{AIC}
fails to decrease.

%% However, there is a snag with this simple approach:
%% whenever we add a new expression $e$ to the DSL,
%% the programs found during waking
%% are not written in terms of $e$ ---
%% and so we must \emph{refactor}
%% the programs in terms of the new DSL component.
%% Concretely, imagine we wanted to discover a new DSL procedure for doubling numbers,
%% after having found the programs
%% \code{(cons (+ 9 9) nil)} and \code{($\lambda$ (x) (+ (car x) (car x)))}.
%% As human programmers,
%% we can look at these pieces of code and recognize that,
%% if we define a new procedure called \code{double},
%% defined as \code{($\lambda$ (x) (+ x x))},
%% then we can rewrite the original programs as
%% \code{(cons (double 9) nil)} and \code{($\lambda$ (x) (double (car x)))}.
%% This process is a kind of refactoring
%% where a new subroutine is defined (\code{double})
%% and the old programs rewritten
%% in terms of the new subroutine.
%% Figure~\ref{mapFactor}A diagrams a refactoring process where the agent
%% must rediscover the higher-order function \code{map} starting from the
%% basics of Lisp and the Y-combinator.

%% We refine our objective to compress
%% \emph{refactorings} of the programs found during waking, minimizing
%% \begin{equation}
%%         -\log \probability[\mathcal{D}] + 
%% \min_{\theta}\Bigg(-\log \probability[\theta|\mathcal{D}] + \|\theta\|_0 +\sum_{x\in X} - \log \underbrace{\sum_{\substack{p:\\\exists p'\in \mathcal{B}_x: p\manyReduce p'}}}_{\text{Refactors }\mathcal{B}_x}\probability[x|p]\probability[p|\mathcal{D},\theta]\Bigg)
%% \label{factorObjective}
%% \end{equation}
%% where $p\manyReduce p'$ is the standard notation for ``expression $p$
%% evaluates to $p'$ by the rules of $\lambda$-calculus''~\cite{pierce}.
%% Equation~\ref{factorObjective} captures the idea that we want to add
%% new components to the DSL while jointly refactoring our old programs
%% in terms of these new components.  But this joint optimization is
%% intractable, because there are infinitely many ways of refactoring a
%% program.  To make refactoring tractable we first limit the degree to
%% which a piece of code can be refactored: rather than consider every
%% refactoring, we bound the number of $\lambda$-calculus evaluation
%% steps separating a refactoring from its original program.  Now the
%% number of refactoring this is finite but astronomically large: for the
%% example in Figure~\ref{mapFactor} there are approximately $10^{14}$
%% possible refactorings -- a quantity that grows exponentially both as a
%% function of program size and a function of the bound on evaluation
%% steps. How can we tame this combinatorial explosion?

@@

  \end{tikzpicture}
  \caption{DSL learning as code refactoring. \textbf{Panel A:} During waking we discover programs for each task, then refactor the code from those programs to expose common subprograms (highlighted in \orange{orange}). Common subprograms are incorporated into the DSL when they increase a Bayesian objective. Intuitively, these new DSL components best compress the programs found during waking. \textbf{Panel B:} \# of possible refactorings grows exponentially with program size, so we represent refactorings using version spaces, which augment syntax trees with a \emph{union} operator whose children are themselves version spaces. Right graph: version spaces are exponentially more efficient than explicitly constructing set of refactorings. In this graph, refactored programs are of the form $1+1+\cdots  + 1$.}\label{mapFactor}
\end{figure*}

@@

Formally, the recognition model $Q(p|x)$ should approximate the posterior
$\probability[p|\mathcal{D},\theta,x]$.
We can either train $Q$ to perform full posterior inference by minimizing the expected KL-divergence, $  \expect\left[\text{KL}\left(\probability[p|x,\mathcal{D},\theta]\|Q(p|x) \right) \right]$,
or we can train $Q$ to perform MAP inference
by maximizing $\expect\left[\max_{p\text{ maxing }\probability[\cdot |x,\mathcal{D},\theta]} \log Q(p|x) \right]$,
where in both cases the expectation is taken over tasks. Taking this expectation over the empirical distribution of tasks trains $Q$ on replays; taking it over samples from the generative model trains $Q$ on fantasies.
We define a pair of alternative objectives for the recognition model,
$\mathcal{L}^{\text{posterior}}$ and $\mathcal{L}^{\text{MAP}}$,
which either train $Q$ to perform full posterior inference or MAP inference, respectively.
These objectives combine replays and fantasies:
\begin{align*}
  \mathcal{L}^{\text{posterior}} &= \mathcal{L}_{\text{Replay}}^{\text{posterior}} + \mathcal{L}_{\text{Fantasy}}^{\text{posterior}}&
  \mathcal{L}^{\text{MAP}} &= \mathcal{L}_{\text{Replay}}^{\text{MAP}} + \mathcal{L}_{\text{Fantasy}}^{\text{MAP}}\\
  \mathcal{L}_{\text{Replay}}^{\text{posterior}}& = \expect_{x\sim X}\left[\sum_{p\in \mathcal{B}_x}
    \frac{\probability\left[x,p|\mathcal{D},\theta \right]\log Q(p|x)}{\sum_{p'\in \mathcal{B}_x}\probability\left[x,p'|\mathcal{D},\theta \right]}\right] &
  \mathcal{L}_{\text{Replay}}^{\text{MAP}}& = \expect_{x\sim X}\left[\max_{\substack{p\in \mathcal{B}_x\\p\text{ maxing }\probability[\cdot |x,\mathcal{D},\theta]}} \log Q(p|x) \right]  \\
  \mathcal{L}_{\text{Fantasy}}^{\text{posterior}} &= \expect_{(p,x)\sim(\mathcal{D},\theta) }\left[\log Q(p|x)\right]&
  \mathcal{L}_{\text{Fantasy}}^{\text{MAP}} &= \expect_{x\sim(\mathcal{D},\theta) }\left[\max_{\substack{p\\p\text{ maxing }\probability[\cdot |x,\mathcal{D},\theta]}}\log Q(p)\right]
\end{align*}
We maximize $\mathcal{L}^{\text{MAP}}$ rather than
$\mathcal{L}^{\text{posterior}}$ for two reasons:
$\mathcal{L}^{\text{MAP}}$ prioritizes the shortest program solving a
task, thus more strongly accelerating enumerative search during waking;
and, combined with our parameterization of $Q$, described next, we
will show that $\mathcal{L}^{\text{MAP}}$ forces the recognition model
to break symmetries in the space of programs.

\noindent\textbf{Parameterizing $Q$.} The recognition model predicts a
fixed-dimensional tensor encoding a distribution over subroutines in
the DSL, conditioned on the local context in the syntax tree of the
program. This local context consists of the parent node in the syntax
tree, as well as which argument is being generated, functioning as a
kind of `bigram' model over trees. Figure~\ref{symmetry} (left)
diagrams this generative process.  This parameterization confers three
main advantages: (1) it supports fast enumeration and sampling of
programs, because the recognition model only runs once per
task, like in~\cite{balog2016deepcoder,ecc,menon2013machine} -- thus
we can fall back on fast enumeration if the target program is unlike
the training programs; (2) the recognition model provides fine-grained
information about the structure of the target program, similar
to~\cite{devlin2017robustfill,zavershynskyi2018naps}; and (3) in
conjunction with $\mathcal{L}^{\text{MAP}}$ the recognition model
learns to break symmetries in the space of programs.

\noindent\textbf{Symmetry breaking.} A good DSL not only exposes high-level building blocks, but also
carefully restricts the ways in which those building blocks are
allowed to compose.  For example,
%% a DSL for list manipulation should
%% contain both the empty list and a routine for appending lists, but
%% should not allow appending the empty list.  Similarly 
a DSL for arithmetic should disallow adding zero, or force
right-associative addition.  These restrictions break symmetries in
the space of programs.  A bigram parameterization of the recognition
model, combined with the $\mathcal{L}^{\text{MAP}}$ training
objective, interact in a way that breaks symmetries in the program
space, allowing the agent to more efficiently explore the space of
programs.  This interaction occurs because the bigram parameterization
can disallow DSL primitives depending on their local syntactic
context, while the $\mathcal{L}^{\text{MAP}}$ objective forces all
probability mass onto a single member of a set of syntactically
distinct but semantically equivalent expressions
(Appendix~\ref{recognitionAppendix}).
We experimentally confirm this symmetry-breaking
behavior by training recognition models that minimize either
$\mathcal{L}^{\text{MAP}}$/$\mathcal{L}^\text{posterior}$ and which
use either a bigram parameterization/unigram\footnote{In the unigram variant $Q$ predicts a $|\mathcal{D}| + 1$-dimensional vector: $Q(p|x) = \probability[p|\mathcal{D},\theta_i = Q_i(x)]$,
  and was used in our prior work~\cite{ecc}} parameterization.
Figure~\ref{symmetry} (right) shows the result of training $Q$ in these four regimes
and then sampling programs.
On this particular run,
the combination of
bigrams and $\mathcal{L}^{\text{MAP}}$ learns to
avoid adding zero and associate addition to the right ---
different random initializations
lead to either right or left association.

\begin{comment}

@@

We adopt a middle ground between these two extremes.  Our recognition
model predicts a distribution over primitives in the DSL,
conditioned on the local context in the syntax tree of the
program. When predicting the next node to add to the syntax tree of a program,
the recognition model
conditions on the parent node, as well as
which argument is being generated.
This is a kind of `bigram' model over trees,
where the bigrams
take the form of (parent, child, argument index).
Figure~\ref{symmetry} (left) diagrams this generative process and Algorithm~\ref{recognitionSample}
specifies a sampling procedure for $Q(\cdot |x)$.
This parameterization
confers three main advantages:
(1) it supports fast enumeration
and sampling of programs,
because the
recognition model
only needs to run once for each task;
(2) it allows the recognition model to provide fine-grained
information about the structure of the target program;
and (3)
training this recognition model
causes it to learn to break symmetries in the space of programs,
described next.

@@

\subsection{Programs that manipulate sequences}\label{sequences}
We first apply \system to two classic benchmark domains: list
processing and text editing. In both cases we solve tasks specified by
a input/output examples, starting with a generic functional programming
basis: \code{foldr}, \code{unfold}, \code{if}, \code{map},
\code{length}, \code{index}, \code{=}, \code{+}, \code{-}, \code{0},
\code{1}, \code{cons}, \code{car}, \code{cdr}, \code{nil}, and
\code{is-nil}.
